---
title: "Analysis of Killings by US Police"
author: "Trevor McGlynn"
date: 2020-06-24
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(quantreg)

PALLETTE <- function(){
  return(scale_fill_brewer(palette = "Reds"))
}
police <- read.csv('PoliceKillingsUS.csv')
```
## Dataset for Police Killings
This first dataset contains observations for police killings which occurred in the US. It's variables are a name, date, and ID for each case, the city and state where the incident occurred, the victim's age and race, and some other details about the victim's encounter with police (such as whether they had a weapon, or whether the police officer was wearing a body cam). 

```{r polive-overview, echo=FALSE}
unique(police$manner_of_death)
sample(unique(police$armed), 20)
unique(police$race)
sample(police$city, 20)
unique(police$threat_level)
unique(police$flee)
```
From this set, we can make some selections on what data will be pertinent to our analysis. Since females account for only `r (1- sum(police$gender == 'M') / nrow(police)) * 100`% of the total observations we'll be leaving that column out. Further, looking at the variables `threat_level` and `manner_of_death`, neither seem to determine anything useful. Variable 'threat_level' contains 
factors (attack, other, undetermined) that don't sound reliable, or meaningful: 2 of 3 are ambiguous. Variable 'manner_of_death' contains factors `unique(police$manner_of_death)`. For an analysis purely interested in fatalities it seems inconsequential to look at what other injury occurred and just be looking at the nature of the fatality and some other statistical data related to the environment where it happened.

That said, the scope of this analysis will look at the following from this table:
* Factors during police encounter (race, weapons, mental illness)
* City and state
* Body camera usage
* Dates (full annual data for 2015 and 2016, partial 2017)
```{r pipe-police-tbl}
# This pipe creates tibble 'df' from 'police' with column selections.
# Adds: year, month, day, unjust_force
# 'unjust_force': Let's suppose justified use of force would be when an officer 
# encounters someone who has equal power of force (another gun). We'll look at 
# those in a different context. The use of force is arguably more nuanced than 
# this, however it creates a good baseline to frame what is, and is not, 
# 'appropriate' use of deadly force by equally matching force available.

police <- police %>%
  filter(gender == 'M') %>%
  select(-gender, -manner_of_death, -threat_level) %>%
  rename(illness = signs_of_mental_illness, cam = body_camera) %>%
  mutate(date = dmy(as.character(date)),
         year = year(date),
         month = month(date),
         day = day(date),
         name = as.character(name),
         unjust_force = armed == 'unarmed') %>%
  select(-date) %>%
  arrange(year, month, day)
```
## Dataset for Demographic Data
This next data set joins demographics for cities across the US. Statistics included are the following:
* Median Income for 2015
* Percentage of People Below Poverty Level
* Percent of Population > 25 Who Completed High School
* Share of Race by City

These will be joined with a cities table based on city and state. The `cities` table is derived from the `police` table and will include all cities where there was a victim from a police shooting, and a count of cases for that particular city. We'll also tidy the data for upcoming statistical summaries. 
```{r demographics-setup, message=FALSE, warning=FALSE}
# Create a 'cities' data frame from 'df' which will hold all unique city names 
# and their count of police killings, arranged by state then city, which seemed
# like the simplest way to match the tables quickly, given the data on hand
cities <- police %>%
  group_by(city, state) %>%
  select(-name) %>%
  summarize(deaths = n()) %>%
  arrange(state, city)

# Metro statistical data here
stats <- read.csv('MedianHouseholdIncome2015.csv') %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City)
stats <- left_join(stats, read.csv('PercentagePeopleBelowPovertyLevel.csv') %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City))
stats <- left_join(stats, read.csv('PercentOver25CompletedHighSchool.csv') %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City))
stats <- left_join(stats, read.csv('ShareRaceByCity.csv') %>%
  rename(Geographic.Area = Geographic.area) %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City))

# Cleans up some column names and runs a regular expression through the city
# names to make match against 'df' containing the police killings data. 
# Looks like it's statistical classification was appended in each table?
stats <- stats %>% 
  rename(state = Geographic.Area, 
         city = City, 
         income = Median.Income,
         hs_completion = percent_completed_hs, 
         W = share_white, 
         B = share_black,
         N = share_native_american, 
         A = share_asian, 
         H = share_hispanic,
         poverty = poverty_rate) %>%
  mutate(city = sub('(.*)(\\s\\w+)$', '\\1', city))

cities <- left_join(cities, stats)
```
```{r demographic-tidy, echo=FALSE, message=FALSE, warning=FALSE}
cities$income <- (as.numeric(levels(cities$income))[cities$income])
cities$poverty <- (as.numeric(levels(cities$poverty))[cities$poverty]) / 100
cities$hs_completion <- (as.numeric(levels(cities$hs_completion))[cities$hs_completion]) / 100
cities$W <- (as.numeric(levels(cities$W))[cities$W]) / 100
cities$B <- (as.numeric(levels(cities$B))[cities$B]) / 100
cities$N <- (as.numeric(levels(cities$N))[cities$N]) / 100
cities$A <- (as.numeric(levels(cities$A))[cities$A]) / 100
cities$H <- (as.numeric(levels(cities$H))[cities$H]) / 100
rm(stats)
cities <- cities %>% ungroup(city, state)
head(cities)
```
# Univariate Plots
This is a preliminary exploration of the dataset for summaries and univariate plots.

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

## Univariate Summaries
```{r echo=FALSE, Univariate-summaries}
table_styles = c('striped', 'hover', 'condensed', 'responsive')
# Weapons 
(weapons_uni <- police %>% select(armed) %>% group_by(armed) %>% count(sort = TRUE))
# Victim stats
victims_uni <- police %>% select(age, race, illness, year:day)
kable(summary(victims_uni), caption = 'Statistics about Victims') %>% 
  kable_styling(bootstrap_options = table_styles)
# Police stats
police_uni <- police %>% select(city, state, flee:unjust_force)
kable(summary(police_uni), caption = 'Summary Statistics about Police') %>% 
  kable_styling(bootstrap_options = table_styles)

# Demographics except race    
dems_uni <- cities %>% select(-city, -(W:H))
kable(summary(dems_uni), caption = 'Demographics for Represented Cities') %>%
  kable_styling(bootstrap_options = table_styles)

# Racial Makeup of rep cities
(race_uni <- cities %>% 
  select(W:H) %>%
  summarize_all(funs(mean), na.rm = TRUE))
```
## Univariate Plots
```{r Univariate-plots, echo=FALSE}
## WEAPONS plotting
ggplot(weapons_uni %>% group_by(armed) %>% filter(n > 9) %>% arrange(desc(n)), 
       mapping = aes(reorder(armed, -n), n, fill = armed)) +
  geom_col() +
  scale_y_log10() +
  coord_flip() +
  labs(title = 'Top Weapons Used', y = 'Count (logarithmic)', x = 'Weapon Used') +
  scale_fill_brewer(palette = 'Blues')

## VICTIMS plotting
ggplot(victims_uni %>% select(age, race),
       mapping = aes(x = race, y = age, fill = race)) +
  geom_boxplot() +
  scale_fill_brewer(palette = 'Blues') +
  scale_y_continuous(breaks = seq(10, 100, by = 10)) +
  labs(title = 'Age Distribution by Race', x = 'Race', y = 'Age')

ggplot(victims_uni %>% select(age), mapping = aes(x = age, fill = 'red')) +
  geom_dotplot(binwidth = 0.5, show.legend = FALSE) +
  labs(title = 'Distribution of Ages of Victims', x = 'Age')

## DEMOGRAPHICS plotting
ggplot(dems_uni %>% select(income), mapping = aes(income, fill = 'red')) +
  geom_dotplot(binwidth = 1400, show.legend = FALSE) +
  labs(title = 'Represented Cities: Median Income', x='Median Income', y='')

ggplot(dems_uni %>% select(poverty), mapping = aes(poverty, fill = 'red')) +
  geom_dotplot(binwidth = 0.008, method = 'histodot', show.legend = FALSE) +
  labs(title = 'Represented Cities: Poverty Rate', x='Poverty Rate', y='')

ggplot(dems_uni %>% select(hs_completion), mapping = aes(hs_completion, fill = 'red')) +
  geom_dotplot(binwidth = 0.008, show.legend = FALSE) +
  labs(title = 'Represented Cities: HS Graduation Rate', x='HS Graduation Rate', y='')
```
# Univariate Analysis

## Age of Victims
The median age for victims of police shootings is `r summary(victims_uni$age)['Median']`, which is higher than expected, and leaving the data set skewed right according to age.

## Demographics Data for Cities
**Median income** across the represented cities is `r summary(dems_uni$income)['Median']`
**Poverty rates** across the represented cities average `r summary(dems_uni$poverty)['Mean']`
**High School Completion Rates** across the represented cities average `r summary(dems_uni$hs_completion)['Mean']`

## Weapons Used
Guns were the overwhelming majority of weapons used by victims during encounters where police shootings occur, followed by knives and vehicles. Some of the more rare weapons can be seen below.

With weapons, one issue seems to be duplicate categories for unidentified weapons: `undetermined` and `unknown_weapon`. This must be address before further comparison.

```{r interesting-weapons, echo=FALSE}
weapons_uni %>% filter(n == 1)
rm(police_uni, victims_uni, weapons_uni)
```
# Bivariate Plots

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r Bivariate-plots, echo=FALSE}
## Demographics Plotting
ggplot(cities %>% select(deaths, income) %>% filter(deaths > 1),
       mapping = aes(income, deaths, color = 'red')) +
  geom_jitter(show.legend = FALSE) +
  scale_x_continuous(limits = c(35000, 60000))
ggplot(cities %>% select(deaths, poverty),
       mapping = aes(poverty, deaths, color = 'red')) +
  geom_jitter(show.legend = FALSE)
ggplot(cities %>% select(deaths, hs_completion) %>% filter(deaths > 1),
       aes(hs_completion, deaths, color = 'red')) + geom_jitter()

## Use of Force by Age
ggplot(police %>% 
         filter(year != 2017) %>% 
         select(age, race, unjust_force) %>%
         mutate(poc = race %in% c('B', 'H', 'N')) %>%
         filter(poc == TRUE | race == 'W'),
       aes(poc, age, fill = poc)) +
  geom_violin(scale = 'area') +
  facet_wrap(~ unjust_force) +
  scale_fill_brewer(palette = 'Blues') +
  labs(title = 'Use of Unjust Force', 
       x = 'Person of Color (T/F)', y = 'Age')

## Count by month, by race
ggplot(police %>% filter(year != 2017) %>% select(month, race), 
       aes(month, fill = race)) +
  geom_bar(position = 'fill') +
  scale_x_continuous(limits = c(0, 12)) +
  scale_fill_brewer(palette = 'Blues') +
  labs(title = 'Proportion of Deaths by Race and Month',
       x = 'Month', y = 'Propotion of Deaths')

```
# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Relationships

#### Demographic Data
Surprisingly, there was no clear direct relationship between the various demographics data and increased police killings. Instead, it seems that some municipalities are just far, far worse than others when it comes to victims of police shooting. Trends may be easier to see on a state level; cities are too granular for these correlations. Outliers that have a disproportionate amount of deaths typically cluster around the mean for all three metrics (income, poverty, and high school completion)

### Use of Force
Proportionally against all cases of police killings, people of color face disproportionate unjust use of force, where "unjust" refers to a police officer killing an unarmed person.

For whites, use of force peaks in the victim's late twenties. For people of color, peak it in the early twenties, followed by a second peak for victims in their late 30's. This reflects the median age for victims which is `r median(summary(police$age))`.

A cursory glance against the distribution of races represented in police killings shows the disproportion of violence directed towards people of color against their respective proportions of the population. From this sample, whites represent `r race_uni['W'] * 100`% of the population, whereas people of color represent `r (race_uni['B'] + race_uni['N'] + race_uni['H']) * 100`% of the population. The graph `Proportion of Deaths by Race and Month` shows almost 50/50 representation of the share of killings, white to people of color.

# Multivariate Plots

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
# Body cams, race, and unjust_force
ggplot(police %>% 
         filter(year != 2017) %>%
         select(race, cam, unjust_force),
       aes(race, fill = unjust_force)) +
  geom_bar(position = 'fill') +
  facet_wrap(~ cam) +
  PALLETTE() +
  labs(title = 'Camera Use & Unjust Force by Race (Proportional)',
       x = 'Race', y = 'Prop')

## VICTIMS Plotting: race, age (bucketed), poc, and illness
pipe <- police %>%
  mutate(poc = race %in% c('B','H','N')) %>%
  select(race, age, poc, illness, unjust_force, flee) %>%
  filter(race == 'W' | poc == TRUE) %>%
  drop_na() %>%
  filter(flee != '')
BREAKS = c(0, 18, 25, 35, 45, 65, 100)

ggplot(pipe, mapping = aes(x = cut(age, breaks = BREAKS), fill = poc)) +
  geom_bar(position = 'fill') +
  scale_fill_brewer(palette = 'Blues') +
  facet_wrap(~ unjust_force) +
  labs(title = 'Proportion of Unjust Force Used by Skin Color (Age Groups)',
       x = 'Age Group', y = 'Prop')
```

# Multivariate Analysis

### Relationships

### Interactions Among Features

### OPTIONAL: Models

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(pipe, mapping = aes(x = cut(age, breaks = BREAKS), fill = poc)) +
  geom_bar(position = 'fill') +
  scale_fill_brewer(palette = 'Blues') +
  facet_wrap(~ unjust_force) +
  labs(title = 'Proportion of Unjust Force Used by Skin Color (Age Groups)',
       x = 'Age Group', y = 'Prop')
```

### Description One
This first plot shows quite clearly how, by proportion of police-induced fatalities, people of color make up a dominant proportion of cases against nearly all age groups **where unjust force is used**. Unjust force was defined in the context of this analysis as the *use of deadly force by a police officer on an unarmed victim*. These graphs show the proportion of total fatalities by police, however all of this becomes more striking when considering the racial makeup of the sample itself:

  Sample proportion PoC: `r race_uni['B'] + race_uni['N'] + race_uni['H']`
  Sample proportion White: `r as.double(race_uni['W'])`
  
From this analysis, it's clear that people of color are disproportionately represented in unjustified police violence, and are far more likely to encounter it than white counterparts.

### Plot Two
```{r echo=FALSE, Plot_Two}
# Body cams, race, and unjust_force
ggplot(police %>% 
         filter(year != 2017, race != '', race != 'A') %>%
         select(race, cam, unjust_force),
       aes(race, fill = unjust_force)) +
  geom_bar(position = 'fill') +
  facet_wrap(~ cam) +
  scale_fill_brewer(palette = 'Blues') +
  labs(title = 'Camera Use & Unjust Force by Race (Proportional)',
       x = 'Race', y = 'Prop')
```

### Description Two
The race of a suspect is more likely to be recorded as "other" when unjust force is used and the officer is wearing a body camera. Furthermore, the two race factors with the greatest share of unjust force cases is "other" (1st) and "black" (2nd).

### Plot Three
```{r echo=FALSE, Plot_Three}
## Use of Force by Age
ggplot(police %>% 
         filter(year != 2017) %>% 
         select(age, race, unjust_force) %>%
         mutate(poc = race %in% c('B', 'H', 'N')) %>%
         filter(poc == TRUE | race == 'W'),
       aes(poc, age, fill = poc)) +
  geom_violin(scale = 'area') +
  facet_wrap(~ unjust_force) +
  scale_fill_brewer(palette = 'Blues') +
  labs(title = 'Use of Unjust Force', 
       x = 'Person of Color (T/F)', y = 'Age')
```

### Description Three
Unjust force by age and race is again clearly shown through this violin plot, but shows an interesting relationship unjust force against persons of color peaks when the victim is in their early twenties, but then increases again in their late thirties. With George Floyd's death in the spotlight, and many brown men in their late-thirties who have died from unjust use of police force, it's important to consider this and *why* it may be.

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!


