---
title: "Analysis of Killings by US Police"
author: "Trevor McGlynn"
date: 2020-06-24
output: html_document
---
```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(magrittr)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(quantreg)

police <- read.csv('PoliceKillingsUS.csv')
BREAKS <- c(0, 18, 25, 30, 35, 45, 65, 100)
```
# Introduction
This analysis looks at circumstantial observations for all reported cases of death by police officer between 2015 and 2016. It should be worth mentioning that this analysis comes in the wake of nationwide protests following the deaths of George Floyd, Breonna Taylor, and many more as a result of police violence, predominantly against people of color. The data show the disproportional representation of people of color who are killed by police, and frames this around the idea of *unjust force*: the use of deadly force against someone who is completely unarmed. (This warrants more discussion, as will be mentioned shortly.)

Here is a breakdown of the data set used:

```{r echo=FALSE, results='asis'}
files <- c('MedianHouseholdIncome2015.csv','PercentagePeopleBelowPovertyLevel.csv',
          'PercentOver25CompletedHighSchool.csv', 'PoliceKillingsUS.csv',
          'ShareRaceByCity.csv')
sizes <- c('709.0', '664.4', '676.3', '227.7', '1094.4')
kable(tibble(Files = files, Sizes_kB = sizes)) %>%
  kable_styling()
```

Respectively, these represent data from the 2015 census by census area for median income, percentage of the population below the federal poverty level, the percent of the population that are high school graduates over 25 years old, reported instances of police killings between 2015 and 2017 in the US, and the racial makeup of each city (expressed as a percentage).

### Data Frames Used
* `police` <- Observations of individual killings by police officers
* `cities` <- Combined statistical information about cities represented in the `police` DF

### Potential Issues
*Unjust force* <- BOOL, this indicates a victim was completely unarmed when killed by a police officer. However, this does not account for all cases where one could *reasonably conclude force was unjustified*. This is the lowest baseline that could define force in the context of just or unjust.

*Illness, flee* <- These are subjective judgments given by police officers and could be unreliable. 

## Dataset for Police Killings
```{r polive-overview, eval=FALSE, echo=FALSE}
unique(police$manner_of_death)
sample(unique(police$armed), 20)
unique(police$race)
sample(police$city, 20)
unique(police$threat_level)
unique(police$flee)
```
This first data set contains observations for police killings which occurred in the US. Its variables are a name, date, and ID for each case, the city and state where the incident occurred, the victim's age and race, and some other details about the victim's encounter with police (such as whether they had a weapon, or whether the police officer was wearing a body cam). 

Looking at the variables `threat_level` and `manner_of_death`, neither seem to determine anything useful. Variable 'threat_level' contains factors (attack, other, undetermined) that don't sound reliable, or meaningful: 2 of 3 are ambiguous. Variable 'manner_of_death' contains factors `unique(police$manner_of_death)`. For an analysis purely interested in fatalities it seems inconsequential to look at what other injury occurred and just be looking at the nature of the fatality and some other statistical data related to the environment where it happened.

That said, the scope of this analysis will look at the following from this table:

* Factors during police encounter (race, weapons, mental illness)
* City and state
* Body camera usage
* Dates (in 2015 and 2016)
* Victim characteristics (age, gender, and race)

```{r pipe-police, results='hide'}
police <- police %>%
  filter(race != '') %>%
  select(-manner_of_death, -threat_level, flee) %>%
  rename(illness = signs_of_mental_illness, cam = body_camera) %>%
  mutate(name = as.character(name),
         unjust_force = armed == 'unarmed',
         poc = race %in% c('B', 'H', 'N'),
         date = dmy(date),
         age_range = cut(age, breaks = BREAKS)) %>%
  filter(year(date) != 2017) %>%
  arrange(date)
```
## Dataset for Demographic Data
The second data set will join demographics data against cities represented with deaths in the `police` data set by city and state. We'll also tidy the data for upcoming statistical summaries. 
```{r pipe-demographics, message=FALSE, warning=FALSE}
# Create a 'cities' data frame from 'df' which will hold all unique city names 
# and their count of police killings, arranged by state then city, which seemed
# like the simplest way to match the tables quickly, given the data on hand
cities <- police %>%
  group_by(city, state) %>%
  select(-name) %>%
  summarize(deaths = n()) %>%
  arrange(state, city)
```
```{r demographics-tidy, include=FALSE, message=FALSE, warning=FALSE}
# Metro statistical data here
stats <- read.csv('MedianHouseholdIncome2015.csv') %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City)
stats <- left_join(stats, read.csv('PercentagePeopleBelowPovertyLevel.csv') %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City))
stats <- left_join(stats, read.csv('PercentOver25CompletedHighSchool.csv') %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City))
stats <- left_join(stats, read.csv('ShareRaceByCity.csv') %>%
  rename(Geographic.Area = Geographic.area) %>%
  group_by(City) %>% 
  arrange(Geographic.Area, City))

# Cleans up some column names and runs a regular expression through the city
# names to make match against 'df' containing the police killings data. 
# Looks like it's statistical classification was appended in each table?
stats <- stats %>% 
  rename(state = Geographic.Area, 
         city = City, 
         income = Median.Income,
         hs_completion = percent_completed_hs, 
         W = share_white, 
         B = share_black,
         N = share_native_american, 
         A = share_asian, 
         H = share_hispanic,
         poverty = poverty_rate) %>%
  mutate(city = sub('(.*)(\\s\\w+)$', '\\1', city))

cities <- left_join(cities, stats)

cities$income <- (as.numeric(levels(cities$income))[cities$income])
cities$poverty <- (as.numeric(levels(cities$poverty))[cities$poverty]) / 100
cities$hs_completion <- (as.numeric(levels(cities$hs_completion))[cities$hs_completion]) / 100
cities$W <- (as.numeric(levels(cities$W))[cities$W]) / 100
cities$B <- (as.numeric(levels(cities$B))[cities$B]) / 100
cities$N <- (as.numeric(levels(cities$N))[cities$N]) / 100
cities$A <- (as.numeric(levels(cities$A))[cities$A]) / 100
cities$H <- (as.numeric(levels(cities$H))[cities$H]) / 100
rm(stats)
cities <- cities %>% ungroup(city, state)
head(cities)
```

## Univariate Plots
```{r Univariate-plots, echo=FALSE, warning=FALSE}
# Deaths plotted over time
ggplot(police %>% select(date), aes(date)) +
  geom_histogram(binwidth = 7, na.rm = TRUE, fill = 'Midnight Blue', color = 'white') +
  labs(title = 'Killings by Week Across Range', x = 'Date', y = 'Count')

# Weapons involved, not including guns
ggplot(police %>% 
         select(armed) %>% 
         group_by(armed) %>%
         mutate(count = n()) %>%
         filter(between(count, 6, 1000),
                armed != 'undetermined',
                armed != 'unknown weapon'),
       aes(armed)) +
  geom_bar(fill = 'Midnight Blue', color = 'white') +
  labs(title = 'Frequent Weapons Involved', x = 'Weapon', y = 'Count')

# Age distribution of Victims
ggplot(police %>% select(age), aes(age)) +
  geom_dotplot(binwidth = 0.5, show.legend = FALSE, color = 'white',  fill = 'Midnight Blue') +
  labs(title = 'Distribution of Ages of Victims', x = 'Age')

# Income distributions of Represented cities
ggplot(cities %>% select(income), aes(income)) +
  geom_dotplot(binwidth = 1400, show.legend = FALSE, color = 'white',  fill = 'Midnight Blue') +
  labs(title = 'Represented Cities: Median Income', x='Median Income', y='')

# Poverty distributions of Represented cities
ggplot(cities %>% select(poverty), aes(poverty)) +
  geom_dotplot(binwidth = 0.008, show.legend = FALSE, color = 'white',  fill = 'Midnight Blue') +
  labs(title = 'Represented Cities: Poverty Rate', x='Poverty Rate', y='')

# High School Completion distributions of Represented cities
ggplot(cities %>% select(hs_completion), mapping = aes(hs_completion)) +
  geom_dotplot(binwidth = 0.008, show.legend = FALSE, color = 'white',  fill = 'Midnight Blue') +
  labs(title = 'Represented Cities: HS Graduation Rate', x='HS Graduation Rate', y='')
```
## Univariate Summaries
```{r echo=FALSE, Univariate-summaries}
table_styles = c('striped', 'hover', 'condensed', 'responsive')


# Information about victims
armed_stats <- police %>% select(armed) %>%
  filter(armed != 'undetermined') %>%
  filter(armed != 'unknown weapon')
summary(armed_stats)
force_stats <- police %>% select(unjust_force, illness, cam)
summary(force_stats)
city_state_stats <- police %>% select(city, state)
summary(city_state_stats)
age_race_stats <- police %>% select(age, race, poc)
summary(age_race_stats)

# Demographics on deaths, income, poverty levels, and HS completion   
dems_stats <- cities %>% 
  group_by(state) %>% ungroup(state) %>%
  select(deaths, income, poverty, hs_completion)
summary(dems_stats)

race_makeup <- cities %>% select(W:H)
summary(race_makeup)
```
## Univariate Analysis

#### Age of Victims
The median age for victims of police shootings is `r median(summary(age_race_stats$age))`, which is higher than expected, and leaving the data set skewed right according to age.

#### Demographics Data for Cities
**Median income** across the represented cities is `r as.integer(median(summary(dems_stats$income)))`
**Poverty rates** across the represented cities average `r round(mean(summary(dems_stats$poverty)), 4)`
**High School Completion Rates** across the represented cities average `r round(mean(summary(dems_stats$hs_completion)), 4)`

#### Weapons Used
Guns were the overwhelming majority of weapons used by victims during encounters where police shootings occur, followed by knives and vehicles. The next observation is unarmed, which occurred `r round((271 / nrow(police)), 4) * 100`% of the time. Exposed in this is also a factor "toy weapon", which occurred `r round((86 / nrow(police)), 4) * 100`%, where a weapon was found to be a toy instead of an actual firearm.

# Bivariate Plots
```{r Bivariate-plots, echo=FALSE, warning=FALSE, message=FALSE}
## Demographics correlations
state_medians <- cities %>% 
  drop_na() %>%
  group_by(state) %>% 
  summarize(income = median(income), 
            poverty = median(poverty), 
            hs_completion = median(hs_completion))
ggplot(state_medians) +
  geom_jitter(aes(income, poverty)) +
  geom_smooth(aes(income, poverty))
ggplot(state_medians) +
  geom_jitter(aes(income, hs_completion)) +
  geom_smooth(aes(income, hs_completion))
ggplot(state_medians) +
  geom_jitter(aes(poverty, hs_completion)) +
  geom_smooth(aes(poverty, hs_completion))

## Age Distribution by Race, BOXPLOT
ggplot(age_race_stats) +
  geom_boxplot(mapping = aes(x = race, y = age, fill = race)) +
  scale_fill_brewer(palette = 'Blues') +
  scale_y_continuous(breaks = seq(10, 100, by = 10))

## Use of Force by Age, VIOLIN PLOT
(p3 <- ggplot(police %>% 
         select(poc, age, race, unjust_force) %>%
         drop_na(),
       aes(poc, age, fill = poc)) +
  geom_violin(scale = 'area') +
  facet_wrap(~ unjust_force) +
  scale_fill_brewer(palette = 'Blues') +
  labs(title = 'Use of Unjust Force', 
       x = 'Person of Color (T/F)', y = 'Age'))

ggplot(police %>% select(unjust_force, poc), aes(poc, fill = unjust_force)) +
  geom_bar(position = 'fill') +
  scale_fill_brewer(palette = 'Blues') +
  scale_y_continuous(limits = c(0, 0.2))

```

# Bivariate Analysis

#### Demographic Data
Correlations between the components of the demographics data by state yielded less than hoped. The was a negative correlation between poverty level and income, naturally, while high school graduation had a weak positive correlation with income. Likewise, high school graduation and poverty rate had a negative correlation.

### Use of Force
People of color, by proportion of fatal interactions with police, occupy almost twice as many cases as non-PoC. They are **far more likely** to be victims of police violence than others.

Unjust force by age and race is again clearly shown through this violin plot, but shows an interesting relationship unjust force against persons of color peaks when the victim is in their early twenties, but then increases again in their late thirties. With George Floyd's death in the spotlight, and many brown men in their late-thirties who have died from unjust use of police force, it's important to consider this and *why* it may be.

### Race and Age
The violin plot on race and age by use of force shows the distribution of age for victims and whether they were a person of color, faceted against the unjust use of force. One key anamoly will be highlighted later in the report.

# Multivariate Plots
```{r echo=FALSE, Multivariate_Plots, echo=FALSE, warning=FALSE}
# Body cams, race, and unjust_force
(p2 <- ggplot(police %>% 
         select(race, cam, unjust_force) %>%
         filter(race != 'A'), #insignificant,
       aes(race, fill = unjust_force)) +
  geom_bar(position = 'fill') +
  facet_wrap(~ cam) +
  scale_fill_brewer(palette = 'Blues') +
  labs(title = 'Camera Use & Unjust Force by Race (Proportional)',
       x = 'Race', y = 'Prop') +
  scale_y_continuous(limits = c(0, 0.75)))

## VICTIMS Plotting: race, age (bucketed), poc, and illness
pipe <- police %>%
  select(race, age, poc, illness, unjust_force, flee, age_range) %>%
  drop_na() %>%
  filter(flee != '')

(p1 <- ggplot(pipe, mapping = aes(age_range, fill = poc)) +
  geom_bar(position = 'fill') +
  scale_fill_brewer(palette = 'Blues') +
  facet_wrap(~ unjust_force) +
  labs(title = 'Proportion of Unjust Force Used by Skin Color (Age Groups)',
       x = 'Age Group', y = 'Prop'))
p1_var <- cities %>% drop_na() %>% summarize(poc = mean(B+H+N), white = mean(W+A))
```
# Multivariate Analysis

### Relationships
* Camera use and the inclination for race to be marked as 'other'
* Use of unjust force against skin color for all age groups is greater for people of color than whites, and this is true across nearly all of the age groups defined in this analysis

The race of a suspect is more likely to be recorded as "other" when unjust force is used and the officer is wearing a body camera. Furthermore, the two race factors with the greatest share of unjust force cases is "other" (1st) and "black" (2nd).

This first plot shows quite clearly how, by proportion of police-induced fatalities, people of color make up a dominant proportion of cases against nearly all age groups **where unjust force is used**. Unjust force was defined in the context of this analysis as the *use of deadly force by a police officer on an unarmed victim*. These graphs show the proportion of total fatalities by police, however all of this becomes more striking when considering the racial makeup of the sample itself:

* **Share of PoC Population** in cities represented by this sample: `r p1_var['poc'] * 100`%
* **Share of non-PoC Population** in cities represented by this sample: `r p1_var['white'] * 100`%
  
From this analysis, it's clear that people of color are disproportionately represented in unjustified police violence, and are far more likely to encounter it than white counterparts.
